\textls[-0]{\thetitle}

### 2023-03-24 canonicitymodel.tex
Fix typo.
Fix typo on identity type.
BugFix.

Like most logical-relations proofs, we interpret each judgment and inductively
interpret each syntax piece. We are working in an intrinsic setting; thus, even if
we omit contexts for brevity, the syntax piece is still well-typed. However,
unlike most logical-relations proofs, our logical-relations model is proof-relevant,
which is essential for modeling universes
properly~\cite{coquand2018canonicity}.
Our canonicity model for the base MLTT fragment is close to the ones
in \citet{coquand2018canonicity} and \citet{sterling2019algebraic}.
% We will omit the treatment of the universe and the verification of most equality for simplicity (and the treatment of the universe will not affect our model). Interested audience can refer to \citet{kaposi2019gluing, coquand2018canonicity,sterling2019algebraic} for a complete technical treatements.

@@

  \denotesC{\goodType{\Gm}{\top}{}}(\gm)(\gm')(t) &= \{\star \} \\

@@

  % \setc{\star}{\text{ when } t \equiv \eqrefl{\sub{a}{\gm}} \text{ and } \sub{a}{\gm} \equiv \sub{b}{\gm}} %\YZ{What if the conditions do not hold? Empty set?}\EDJreply{Yes. It is a family of set indexed by t (or just a function maps t to a set), and if t is neither of the above, then the function will map to empty set.}
  \denotesC{\goodType{\Gm}{\TyPi{A}{B}}{}}(\gm)(\gm')(t) &= \prod_{\goodTerm{\cdot}{u}{\sub{A}{\gm}}} \ \ \prod_{u' \in \denotesC{A}(\gm)(\gm')(u)} \denotesC{B}(\SubstExt{\gm}{u})(\gm', u')(\sub{\app{t}}{\SubstExt{"id"}{u}})\\

@@

  \denotesC{\goodCtx{\cdot}{}}(\gm) &= \{\star\} \quad \text{  a singleton set } \\ 

@@

@@

  \denotesC{\goodTerm{\Gm}{\fst{t}}{T}}(\gm)(\gm') &= \denotesC{t}(\gm)(\gm')[0] \\

@@

\newcommand{\Glued}[1]{\ensuremath{{#1}^\bullet}}
\newcommand{\GluedPi}[2]{\ensuremath{\Pi^\bullet({#1},{#2})}}
\newcommand{\GSubstExt}[2]{\ensuremath{{#1},^\bullet{#2}}}
\newcommand{\Gpair}[2]{\ensuremath{({#1},^\bullet{#2})}}
\newcommand{\Gfst}[1]{\ensuremath{\texttt{fst}^\bullet~{#1}}}
\newcommand{\Gsnd}[1]{\ensuremath{\texttt{snd}^\bullet~{#1}}}
\newcommand{\Gsub}[2]{\ensuremath{{#1}\!\left[{#2}\right]^\bullet}}
\newcommand{\Glam}[1]{\ensuremath{\lambda^\bullet({#1})}}
\newcommand{\Gapp}[1]{\ensuremath{\texttt{app}^\bullet({#1})}}
\newcommand{\GSubstWeak}[1]{\ensuremath{(\texttt{p}^\bullet)^{#1}}}
\newcommand{\GLSigAdd}[3]{\ensuremath{\nu^{+\bullet}({#1},{#2},{#3})}}
\newcommand{\GCaseSig}[3]{\ensuremath{\texttt{CaseTy}^\bullet({#1},{#2},{#3})}}
\newcommand{\Gmodel}[1]{\ensuremath{{({#1})}^c}}
\newcommand{\GEl}[1]{\ensuremath{\Glued{\texttt{El}}({#1})}}
\newcommand{\Gwcode}[1]{\ensuremath{\Glued{\texttt{W}}({#1})}}
\newcommand{\GSubstId}{\ensuremath{\GSubstWeak{0}}}
\newcommand{\Gwsigproj}[3]{\ensuremath{\Glued{{\texttt{w}\pi^{#1}_{\texttt{#2}}}}({#3})}}
\newcommand{\GRecproj}[2]{\ensuremath{\Glued{{\texttt{R}\pi^{#1}}}({#2})}}

@@

  & \text{we define }\TyLkg{}, \TyTkg{} \text{ inductively on its input signature} \\ 

@@

  \denotesC{\goodSig{\Gm}{\LSigAdd{\lsig}{s}{T}}{n+1}} &= (\denotesCC{A}, \denotesCC{s}, \denotesCC{T})::\denotesC{\lsig}\\ 

@@

  \denotesC{\goodTerm{\Gm}{\Tkg{\lkg}}{\TyTkg{\lsig}}} &= \Gmodel{\Gpair{\denotesC{\Tkg{\lkgproj{1}{\lkg}}}}{\Gsub{\Gsub{\denotesC{\lkgproj{2}{\lkg}}}{\GSubstExt{\Glued{\GSubstWeak{1}}}{s}}}{\GSubstExt{\Glued{(\SubstId)}}{\denotesC{\Tkg{\lkgproj{1}{\lkg}}}}}}}

@@

The semantic interpretations above are 
mutually recursively defined with the following inductively defined indexed set $\CWmodel$ (with only one constructor $\CWsup$)
%
\begin{align*}
  "Inductive"~\CWmodel &: (\Glued{\wsig} \in \GWSig{i}{\denotesCC{\cdot}}{N}) \to \{ t : \goodTerm{\cdot}{t}{\El{\wcode{\wsig}}} \} \to \Set{i + 1} \quad "where" \\
    \CWsup &: j < N \to \Glued{a} \in \GTm{\Glued{\cdot}}{\wsigproj{j}{1}{\Glued{\wsig}}} \\
    & \to \Glued{b} \in (\GTm{(\Glued{\cdot} \Glued{,} \Gsub{\wsigproj{j}{2}{\Glued{\wsig}}}{\SubstExt{\GSubstId}{\Glued{a}}}) }{\Gsub{\GEl{\Gwcode{\Glued{\wsig}}}}{\GSubstWeak{1}}}) \to \CWmodel~\Glued{\wsig}~\Wsup{j}{\wsig}{a}{b}
\end{align*}
and its eliminator $\CWrec$
\begin{align*}
  \CWrec \  &: (\Glued{\wsig} \in \GWSig{i}{\denotesCC{\cdot}}{N}) \to 
    (P : \{ t : \goodTerm{\cdot}{t}{\El{\wcode{\wsig}}} \} \to \Set{k})  \\
    & \to [j < N \to  \Glued{a} \in \GTm{\Glued{\cdot}}{\Gwsigproj{j}{1}{\Glued{\wsig}}} \\ 
    & \quad \quad \to 
    \goodTerm{({\cdot} {,} \sub{\wsigproj{j}{2}{{\wsig}}}{\SubstExt{\SubstId}{{a}}}) }{b}{\sub{\El{\wcode{\wsig}}}{\SubstWeak{1}}} \\ 
    & \quad \quad \to b^C : (\Glued{\gm} \in (\Glued{\cdot} \Glued{,} \sub{\Gwsigproj{j}{2}{\Glued{\wsig}}}{\SubstExt{\SubstId}{\Glued{a}}}) \to P~\sub{b}{\Gmodel{\Glued{\gm}}}) \\
    & \quad \quad \to P~(\Wsup{j}{\wsig}{a}{b}) ] \quad \text{ (square bracket here just to make things readable)} \\
    & \to \goodTerm{\cdot}{t}{\El{\wcode{\wsig}}} \to \CWmodel~\Glued{\wsig}~t \to P~t \\
  \CWrec \  & \Glued{\wsig}~P~f~t~(\CWsup~\Glued{a}~\Glued{b}) = f~\Glued{a}~b~(\lambda \Glued{\gm}. \CWrec~P~f~(\sub{b}{\Gmodel{\Glued{\gm}}})~(b^C~\Glued{\gm}))
\end{align*}

Note that in $\CWmodel$, the $\Glued{b}$ uses the definition of
$\denotesC{\wcode{\wsig}}$, which after unfolding, recursively references $\CWmodel$
in a strictly positive position. %So it is a well-founded definition.
We do not distinguish $(b, b^C)$ and $\Glued{b}$ for simplicity. Again, we omit validating the equational rules ($\beta$, $\eta$, and substitution) here.

>>>>

Like most logical-relations proofs, we interpret each judgment and inductively
interpret each syntax piece. We are working in an intrinsic setting; thus, even if
we omit contexts for brevity, the syntax piece is still well-typed. However,
unlike most logical-relations proofs, our logical-relations model is proof-relevant,
which is essential for modeling universes
properly~\cite{coquand2018canonicity}.
Our canonicity model for the base MLTT fragment is close to the ones
in \citet{coquand2018canonicity} and \citet{sterling2019algebraic}, where we utilize the machinary from ambient meta-logic (e.g. use dependent function and dependent tuple in ambient meta-logic to model the one in \TT).
% We will omit the treatment of the universe and the verification of most equality for simplicity (and the treatment of the universe will not affect our model). Interested audience can refer to \citet{kaposi2019gluing, coquand2018canonicity,sterling2019algebraic} for a complete technical treatements.

@@

  \denotesC{\goodType{\Gm}{\top}{}}(\gm)(\gm')(t) &= \{\star \} \quad \text{  a singleton set } \\

@@

  \denotesC{\goodTerm{\Gm}{\Jeq{w}{t}}{\sub{C}{\SubstExt{\SubstExt{id}{v}}{t} }}}(\gm)(\gm') &= \denotesC{w}(\gm)(\gm') \quad \quad \text{ given } \goodTerm{\Gm}{t}{\TyId{u}{v}} \\ 
  & \text{Since $\denotesC{t}(\gm)(\gm')$ witnesses } \sub{t}{\gm} \equiv \eqrefl{\sub{u}{\gm}} \text{ and } \sub{u}{\gm} \equiv \sub{v}{\gm} \\
  \denotesC{\goodType{\Gm}{\TyPi{A}{B}}{}}(\gm)(\gm')(t) &= \prod_{\goodTerm{\cdot}{u}{\sub{A}{\gm}}} \ \ \prod_{u' \in \denotesC{A}(\gm)(\gm')(u)} \denotesC{B}(\SubstExt{\gm}{u})((\gm', u'))(\sub{\app{t}}{\SubstExt{"id"}{u}})\\

@@

  \denotesC{\goodCtx{\cdot}{}}(\gm) &= \{\star\}  \\ 

@@

  \denotesC{\goodSub{\Gm}{\pi_1 \delta}{\Dl}}(\gm)(\gm') &= \denotesC{\delta}(\gm)(\gm')[0] \quad \text{ get the first element of the tuple} \\
  \denotesC{\goodTerm{\Gm}{\pi_2 \delta}{\Dl}}(\gm)(\gm') &= \denotesC{\delta}(\gm)(\gm')[1] \quad \text{ get the second element of the tuple} \\
  \denotesC{\goodSub{\Gm}{\SubstExt{\delta}{t}}{\Dl}}(\gm)(\gm') &= (\denotesC{\delta}(\gm)(\gm'),\denotesC{t}(\gm)(\gm')) \\ 

@@

  \denotesC{\goodTerm{\Gm}{\fst{t}}{T}}(\gm)(\gm') &= \denotesC{t}(\gm)(\gm')[0]  \quad \text{ extract the first element in the tuple }\\

@@

\newcommand{\Glued}[1]{\ensuremath{{#1}^\bullet}}
\newcommand{\GluedPi}[2]{\ensuremath{\Pi^\bullet({#1},{#2})}}
\newcommand{\GSubstExt}[2]{\ensuremath{{#1},^\bullet{#2}}}
\newcommand{\Gpair}[2]{\ensuremath{({#1},^\bullet{#2})}}
\newcommand{\Gfst}[1]{\ensuremath{\texttt{fst}^\bullet~{#1}}}
\newcommand{\Gsnd}[1]{\ensuremath{\texttt{snd}^\bullet~{#1}}}
\newcommand{\Gsub}[2]{\ensuremath{{#1}\!\left[{#2}\right]^\bullet}}
\newcommand{\Glam}[1]{\ensuremath{\lambda^\bullet({#1})}}
\newcommand{\Gapp}[1]{\ensuremath{\texttt{app}^\bullet({#1})}}
\newcommand{\GSubstWeak}[1]{\ensuremath{(\texttt{p}^\bullet)^{#1}}}
\newcommand{\GLSigAdd}[3]{\ensuremath{\nu^{+\bullet}({#1},{#2},{#3})}}
\newcommand{\GCaseSig}[3]{\ensuremath{\texttt{CaseTy}^\bullet({#1},{#2},{#3})}}
\newcommand{\Gmodel}[1]{\ensuremath{{({#1})}^c}}
\newcommand{\GEl}[1]{\ensuremath{\Glued{\texttt{El}}({#1})}}
\newcommand{\Gwcode}[1]{\ensuremath{\Glued{\texttt{W}}({#1})}}
\newcommand{\GSubstId}{\ensuremath{\Glued{"id"}}}
\newcommand{\Gwsigproj}[3]{\ensuremath{\Glued{{\texttt{w}\pi^{#1}_{\texttt{#2}}}}({#3})}}
\newcommand{\GRecproj}[2]{\ensuremath{\Glued{{\texttt{R}\pi^{#1}}}({#2})}}

@@

  \text{we define }\TyLkg{}, \TyTkg{} & \text{ inductively on its input signature, by defining } \mathit{L},\Glued{\mathit{P}} \text{ first}  \\ 

@@

  \denotesC{\goodSig{\Gm}{\LSigAdd{\lsig}{s}{T}}{n+1}} &= (\denotesCC{A}, \denotesCC{s}, \denotesCC{T})::\denotesC{\lsig} \quad \text{ given } \goodTerm{\Gm, \TyTkg{\lsig}}{s}{A} \\ 

@@

  \denotesC{\goodTerm{\Gm}{\Tkg{\lkg}}{\TyTkg{\lsig}}} &= \Gmodel{\Gpair{\denotesC{\Tkg{\lkgproj{1}{\lkg}}}}{\Gsub{\Gsub{\denotesC{\lkgproj{2}{\lkg}}}{\GSubstExt{\Glued{\GSubstWeak{1}}}{s}}}{\GSubstExt{\Glued{("id")}}{\denotesC{\Tkg{\lkgproj{1}{\lkg}}}}}}}

@@

The semantic interpretations above are defined together with the following inductively defined indexed set $\CWmodel$ (with only one constructor $\CWsup$)
%
\begin{align*}
  "Inductive"~\CWmodel &: (\Glued{\wsig} \in \GWSig{i}{\denotesCC{\cdot}}{N}) \to \{ t : \goodTerm{\cdot}{t}{\El{\wcode{\wsig}}} \} \to \Set{i + 1} \quad "where" \\
    \CWsup &: j < N \to \Glued{a} \in \GTm{\Glued{\cdot}}{\wsigproj{j}{1}{\Glued{\wsig}}} \\
    & \to \Glued{b} \in (\GTm{(\Glued{\cdot} \Glued{,} \Gsub{\wsigproj{j}{2}{\Glued{\wsig}}}{\SubstExt{\GSubstId}{\Glued{a}}}) }{\Gsub{\GEl{\Gwcode{\Glued{\wsig}}}}{\GSubstWeak{1}}}) \to \CWmodel~\Glued{\wsig}~\Wsup{j}{\wsig}{a}{b}
\end{align*}
and its eliminator $\CWrec$
\begin{align*}
  \CWrec \  &: (\Glued{\wsig} \in \GWSig{i}{\denotesCC{\cdot}}{N}) \to 
    (P : \{ t : \goodTerm{\cdot}{t}{\El{\wcode{\wsig}}} \} \to \Set{k})  \\
    & \to [j < N \to  \Glued{a} \in \GTm{\Glued{\cdot}}{\Gwsigproj{j}{1}{\Glued{\wsig}}} \\ 
    & \quad \quad \to 
    \goodTerm{({\cdot} {,} \sub{\wsigproj{j}{2}{{\wsig}}}{\SubstExt{"id"}{{a}}}) }{b}{\sub{\El{\wcode{\wsig}}}{\SubstWeak{1}}} \\ 
    & \quad \quad \to b^C : (\Glued{\gm} \in (\Glued{\cdot} \Glued{,} \sub{\Gwsigproj{j}{2}{\Glued{\wsig}}}{\GSubstExt{\GSubstId}{\Glued{a}}}) \to P~\sub{b}{\Gmodel{\Glued{\gm}}}) \\
    & \quad \quad \to P~(\Wsup{j}{\wsig}{a}{b}) ] \quad \text{ (square bracket here just to make things readable)} \\
    & \to \goodTerm{\cdot}{t}{\El{\wcode{\wsig}}} \to \CWmodel~\Glued{\wsig}~t \to P~t \\
  \CWrec \  & \Glued{\wsig}~P~f~t~(\CWsup~\Glued{a}~\Glued{b}) = f~\Glued{a}~b~(\lambda \Glued{\gm}. \CWrec~P~f~(\sub{b}{\Gmodel{\Glued{\gm}}})~(b^C~\Glued{\gm}))
\end{align*}

Note that in $\CWsup$, the $\Glued{b}$ uses the definition of
$\denotesC{\wcode{\wsig}}$, which after unfolding, recursively references $\CWmodel$
in a strictly positive position. %So it is a well-founded definition.
We do not distinguish $(b, b^C)$ and $\Glued{b}$ for simplicity. Again, we omit validating the equational rules ($\beta$, $\eta$, and substitution) here.
